The Rockerverse: Packages and Applications for Containerisation with R

### 2020-03-09 rockerverse.tex
fix multiply defined label
googleCloudRunner on CRAN

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Packages and Applications for Containerisation with R}
\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daroczi, Mark Edmondson, Colin Fay, Ellis Hughes, Lars Kjeldgaard, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Péter Sólymos, Tyson Lee Swetnam, Nitesh Turaga, Charlotte Van Petegem, Jason Williams, Craig Willis, Nan Xiao}

@@

\abstract{%
The Rocker Project provides widely used Docker images for R across
different application scenarios. This articles surveys downstream
projects that build upon the Rocker Project images and presents the
current state of R packages for managing Docker images and controlling
containers. These use cases cover diverse topics such as package
development, reproducible research, collaborative work, cloud-based data
processing, and production deployment of services. The variety of
applications demonstrates the power of the Rocker Project specifically
and containerisation in general. Across the diverse ways to use
containers, we identified common themes: reproducible environments,
scalability and efficiency, and portability across clouds. We conclude
that the current growth and diversification of use cases is likely to
continue its positive impact, but see the need for consolidating the
Rockerverse ecosystem of packages, developing common practices for
applications, and exploring alternative containerisation software.
}

@@

The R community continues to grow. This can be seen in the number of new
packages on CRAN, which is still on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies that are
adopting R, as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. The last few
years, coinciding with the rise of R, have also seen the rise of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distributing and deploying of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides Docker images with R
(see the next section for more details). The considerable uptake and
continued evolution of the Rocker~Project has led to numerous projects
that extend or build upon Rocker images, ranging from
reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.}
research to production deployments. As such, this article presents what
we may call the \emph{Rockerverse} of projects across all development
stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and
environment with other containerisation solutions. Our main
contributions is a coherent picture of the current status of using
containers in, with, and for R.

@@

Docker, an application and service provided by the eponymous company,
has, in just a few short years, risen to prominence for developing,
testing, deploying and distributing computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While related
approaches exist, such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerisation''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operating systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualisation. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, Docker
provides an important scalability opportunity. Beyond this though,
Docker has improved this compartmentalisation by accessing the host
system---generally Linux---through a much thinner and smaller shim than
a full operating system emulation or virtualisation. This
containerisation, also called operating-system-level virtualisation
\citep{wikipedia_contributors_os-level_2020}, makes efficient use of
operating system resources \citep{felter_updated_2015} and allows
another order of magnitude in terms of scalability of deployment
\citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a
whole operating system, a container typically runs only one process. The
single process together with sharing the host's kernel results in a
reduced footprint and faster start times. While Docker makes use of
Linux kernel features, it has become important enough that some required
aspects of running Docker have been added to other operating systems so
that those systems can more efficiently support Docker
\citep{microsoft_linux_2019}. The success of Docker has even paved the
way for industry collaboration and standardisation
\citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the
\textbf{Rocker~Project} has provided a variety of Docker containers
since it began in 2014 \citep{RJ-2017-065}. The Rocker~Project provides
several lines of containers spanning from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for easier maintainability (i.e.,
smaller \texttt{Dockerfile}s), better composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, which provides funding for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software. \textbf{Singularity}
stems from the domain of high-perfomance computing
\citep{kurtzer_singularity_2017} and can also run Docker images. Rocker
images work out of the box if the main process is R, e.g., in
\texttt{rocker/r-base}, but Singularity does not succeed in running
images where there is an init script, e.g., in containers that by
default run RStudio~Server. In the latter case, a \texttt{Singularity}
file, a recipe akin to a \texttt{Dockerfile}, needs to be used to make
necessary adjustments. To date, no comparable image stack to the Rocker
Project's images exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which can
also build \texttt{Dockerfile}s and run Docker images. Proof of concepts
exists for using podman to build and run Rocker
containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}},
but the prevelance of Docker, especially in the broader user community
beyond experts or niche systems and the vast amount of blog posts and
courses for Docker currently cap specific development efforts for both
Singularity and podman in the R community. This might quickly change if
the usability and spread of Singularity or podman increase, or if
security features such as rootless/unprivileged containers, which both
these tools support out of the box, become more sought after.

@@

\begin{table}
\centering
\begin{tabular}{lllllll}
  \toprule
\rotatebox{-90}{Functionality} & \rotatebox{-90}{AzureContainers} & \rotatebox{-90}{babelwhale} & \rotatebox{-90}{dockermachine} & \rotatebox{-90}{dockyard} & \rotatebox{-90}{harbor} & \rotatebox{-90}{stevedore }\\ 
  \midrule
Generate a Dockerfile &  &  &  & \checkmark &  &  \\ 
   \midrule
Build an image & \checkmark &  &  & \checkmark &  &  \\ 
   \midrule
Execute a container locally or remotely & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\ 
   \midrule
Deploy or manage instances in the cloud & \checkmark &  & \checkmark &  & \checkmark & \checkmark \\ 
   \midrule
Interact with an instance (e.g., file transfer) &  & \checkmark & \checkmark &  &  & \checkmark \\ 
   \midrule
Manage storage of images &  &  &  &  & \checkmark & \checkmark \\ 
   \midrule
Supports Docker and Singularity &  & \checkmark &  &  &  &  \\ 
   \midrule
Direct access to Docker API instead of using the CLI &  &  &  &  &  & \checkmark \\ 
   \midrule
Installing Docker software &  &  & \checkmark &  &  &  \\ 
   \bottomrule
\end{tabular}
\caption{R packages with Docker client functionality.} 
\label{tab:clients}
\end{table}

\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}) is no longer
actively maintained, but it should be honorably mentioned as the first R
package for managing Docker images and containers. It uses the
\CRANpkg{sys} package \citep{cran_sys} to run system commands against
the Docker CLI, both locally and through an SSH connection, and it has
convenience functions, e.g., for listing and removing containers/images
and for accessing logs. The outputs of container executions are
converted to appropriate R types. The Docker CLI's basic functionality,
although it evolves quickly and with little concern for avoiding
breaking changes, has remained unchanged in core functions, meaning that
a core function such as
\code{harbor::docker\_run(image = "hello-world")} still works depsite
its stopped development.

@@

\textbf{\pkg{dockermachine}}
(\url{https://github.com/cboettig/dockermachine}) is an R package to
provide a convenient interface to
\href{https://docs.docker.com/machine/overview/}{Docker~Machine} from R.
The CLI tool \texttt{docker-machine} allows users to create and manage a
virtual host on local computers, local data centers, or at cloud
providers. A local Docker installation can be configured to transparntly
forward all commands issued on the local Docker CLI to a selected
(remote) virtual host. Docker~Machine was especially crucial for local
use in the early days of Docker, when no native support was available
for Mac or Windows computers, but it remains relevant for provisioning
on remote systems. The package has not received any updates for two
years, but it is functional with a current version of
\texttt{docker-machine} (\texttt{0.16.2}). It potentially lowers the
barriers for R users to run containers on various hosts if they perceive
that using the Docker~Machine CLI directly as a barrier and it enables
scripted workflows with remote processing.

@@

\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an open
source, open development project for the analysis and comprehension of
genomic data \citep{gentleman_bioconductor_2004}. As of October 30th
2019, the project consists of 1823 R software packages, as well as
packages containing annotation or experiment data. \emph{Bioconductor}
has a semi-annual release cycle, where each release is associated with a
particular version of R, and Docker images are provided for current and
past versions of \emph{Bioconductor} for convenience and
reproducibility. All images, which are described on the
\emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), are created with
\texttt{Dockerfile}s maintained on GitHub and distributed through
Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
\emph{Bioconductor}'s ``base'' Docker images are built on top of the
\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
based on the R version in combination with the Bioconductor version and,
therefore, uses Bioconductor version tagging \texttt{devel} and
\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
combinations of R and \emph{Bioconductor} will therefore be accessible
via a specific image tags.

The \emph{Bioconductor} \texttt{Dockerfile} selects the desired R
version from Rocker images, adds required system dependencies, and uses
the \CRANpkg{BiocManager} package for installing appropriate versions of
\emph{Bioconductor} packages \citep{cran_biocmanager}. A strength of
this approach is that the responsibility for complex software
configuration and customization is shifted from the user to the
experienced \emph{Bioconductor} core team. However, a recent audit of
the \emph{Bioconductor} image stack \texttt{Dockerfile} led to the
deprecation of several community-maintained images, because the numerous
specific images became too hard to understand, complex to maintain, and
cumbersome to customise. As part of the simplification, a recent
innovation is the \texttt{bioconductor\_docker:devel} image, which
emulates the \emph{Bioconductor} environment for nightly builds as
closely as possible. This image contains the environment variables and
the system dependencies needed to install and check almost all
\emph{Bioconductor} software packages (1813 out of 1823). It saves users
and package developers from creating this environment themselves.
Furthermore, the image is configured so that \code{.libPaths()} has
\samp{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can
persistently manage installed packages across Docker containers or image
updates. Many R users pursue flexible work flows tailored to particular
analysis needs rather than standardized work flows. The new
\texttt{bioconductor\_docker} image is well suited for this preference,
while \texttt{bioconductor\_docker:devel} provides developers with a
test environment close to \emph{Bioconductor}'s build system.

\label{datascience} \textbf{Data science} is a widely discussed topic in
all academic disciplines \citep[e.g.,][]{donoho_50_2017}. These
discussions have shed a light on the tools and craftspersonship behind
the analysis of data with computational methods. The practice of data
science often involves combining tools and software stacks and requires
a cross-cutting skillset. This complexity and an inherent concern for
openness and reproducibility in the data science community has led to
Docker being used widely. The remainder of this section presents example
Docker images and image stacks featuring R intended for data science.

@@

  the existing front-end configuration, but also to provide a consistent

@@

  management,\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.

@@

\textbf{\pkg{containerit}}
(\url{https://github.com/o2r-project/containerit/}) attempts to take
this one step further and includes these tools to automatically create a
\texttt{Dockerfile} that can execute a given workflow
\citep{nust_containerit_2019}. \pkg{containerit} accepts an R object of
classes \code{"sessionInfo"} or \code{"session\_info"} as input and
provides helper functions to derive these from workflows, e.g., an R
script or R Markdown document, by analysing the session state at the end
of the workflow. It relies on the \pkg{sysreqs}
(\url{https://github.com/r-hub/sysreqs/}) package and it's mapping of
package system dependencies to platform specific installation package
names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
\pkg{containerit} uses \pkg{stevedore} to streamline the user
interaction and improve the created \texttt{Dockerfile}s, e.g., by
running a container for the desired base image to extract the already
available R packages.

@@

\textbf{\CRANpkg{liftr}} \citep{liftr2019} aims to solve the problem of
persistent reproducible reporting in statistical computing based on the
R Markdown format \citep{xie2018}. The irreproducibility of authoring
environments can become an issue for collaborative documents and
large-scale platforms for processing documents. \pkg{liftr} makes the
dynamic R Markdown document the main and sole workflow control file and
the only file that needs to be shared between collaborators for
consistent environments, e.g.~demonstrated in the DockFlow project
(\url{https://dockflow.org}). It introduces new fields to the document
header, allowing users to manually declare the versioned dependencies
required for rendering the document. The package then generates a
\texttt{Dockerfile} from this metadata and provides a utitility function
to render the document inside a Docker container, i.e.,
\code{render\_docker("foo.Rmd")}. An RStudio addin even allows
compilation of documents with the single push of a button.

System dependencies are the domain of Docker, but for a full description
of the computing environment one must also manage the R version and the
the R packages. R versions are available via the versioned Rocker image
stack \citep{RJ-2017-065}.
\href{https://github.com/ColinFay/ronline}{r-online} leverages these
images and provides an app for helping users to detect breaking changes
between different R versions and for historic exploration of R. With a
standalone NodeJS app or
\href{https://srv.colinfay.me/r-online}{r-online}, the user can compare
a piece of code run in two separate versions of R. Internally, r-online
opens one or two Docker instances with the given version of R based on
Rocker images, executes a given piece of code, and returns the result to
the user. Regarding R package management, this can be achieved with
MRAN, or with packages such as \CRANpkg{checkpoint}
\citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can
naturally be applied within images and containers. For example,
\pkg{renv} helps users to manage the state of the R library in a
reproducible way, further providing isolation and portability. While
\pkg{renv} does not cover system dependencies, the \pkg{renv}-based
environment can be transferred into a container either by restoring the
environment based on the main configuration file \texttt{renv.lock} or
by storing the \pkg{renv}-cache on the host and not in the container
\citep{ushey_using_2019}. With both the system dependencies and R
packages consciously managed in a Docker image, users can start using
containers as the \emph{only} environment for their workflows, which
allows them to work indepenently of physical
computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.}
and to assert a specific degree of confidence in the stability of a
developed software
\citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

lines or R

@@

run any container, the package \pkg{googleCloudRunner} demonstrates how
R can be used as the control language for one-time or batch processing
jobs and scheduling of

>>>>

% !TeX root = RJwrapper.tex
\title{The Rockerverse: Packages and Applications for Containerisation with R}
\author{by Daniel Nüst, Dirk Eddelbuettel, Dom Bennett, Robrecht Cannoodt, Dav Clark, Gergely Daróczi, Mark Edmondson, Colin Fay, Ellis Hughes, Lars Kjeldgaard, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Péter Sólymos, Tyson Lee Swetnam, Nitesh Turaga, Charlotte Van Petegem, Jason Williams, Craig Willis, Nan Xiao}

@@

\abstract{%
The Rocker Project provides widely used Docker images for R across
different application scenarios. This article surveys downstream
projects that build upon the Rocker Project images and presents the
current state of R packages for managing Docker images and controlling
containers. These use cases cover diverse topics such as package
development, reproducible research, collaborative work, cloud-based data
processing, and production deployment of services. The variety of
applications demonstrates the power of the Rocker Project specifically
and containerisation in general. Across the diverse ways to use
containers, we identified common themes: reproducible environments,
scalability and efficiency, and portability across clouds. We conclude
that the current growth and diversification of use cases is likely to
continue its positive impact, but see the need for consolidating the
Rockerverse ecosystem of packages, developing common practices for
applications, and exploring alternative containerisation software.
}

@@

The R community continues to grow. This can be seen in the number of new
packages on CRAN, which is still on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies that are
adopting R, as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
These trends cement the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. The last few
years, coinciding with the rise of R, have also seen the rise of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distributing and deploying of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides Docker images with R
(see the next section for more details). The considerable uptake and
continued evolution of the Rocker~Project has led to numerous projects
that extend or build upon Rocker images, ranging from
reproducible\footnote{"Reproducible" in the sense of the \emph{Claerbout/Donoho/Peng} terminology \citep{barba_terminologies_2018}.}
research to production deployments. As such, this article presents what
we may call the \emph{Rockerverse} of projects across all development
stages: early demonstrations, working prototypes, and mature products.
We also introduce related activities that connect the R language and
environment with other containerisation solutions. Our main contribution
is a coherent picture of the current status of using containers in,
with, and for R.

@@

Docker, an application and service provided by the eponymous company,
has, in just a few short years, risen to prominence for developing,
testing, deploying and distributing computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While related
approaches exist, such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synonymous with ``containerisation''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operating systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualisation. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, Docker
provides an important scalability opportunity. Beyond this though,
Docker has improved this compartmentalisation by accessing the host
system---generally Linux---through a much thinner and smaller shim than
a full operating system emulation or virtualisation. This
containerisation, also called operating-system-level virtualisation
\citep{wikipedia_contributors_os-level_2020}, makes efficient use of
operating system resources \citep{felter_updated_2015} and allows
another order of magnitude in terms of scalability of deployment
\citep[cf.][]{datadog_8_2018}, because virtualisation may emulate a
whole operating system, a container typically runs only one process. The
single process together with sharing the host's kernel results in a
reduced footprint and faster start times. While Docker makes use of
Linux kernel features, it has become important enough that some required
aspects of running Docker have been added to other operating systems so
that those systems can more efficiently support Docker
\citep{microsoft_linux_2019}. The success of Docker has even paved the
way for industry collaboration and standardisation
\citep{oci_open_2019}.

@@

For statistical computing and analysis centred around R, the
\textbf{Rocker~Project} has provided a variety of Docker containers
since it began in 2014 \citep{RJ-2017-065}. The Rocker~Project provides
several lines of containers spanning from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for easier maintainability (i.e.,
smaller \texttt{Dockerfile}s), better composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, which provides funding for the project's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software. \textbf{Singularity}
stems from the domain of high-performance computing
\citep{kurtzer_singularity_2017} and can also run Docker images. Rocker
images work out of the box if the main process is R, e.g., in
\texttt{rocker/r-base}, but Singularity does not succeed in running
images where there is an init script, e.g., in containers that by
default run RStudio~Server. In the latter case, a \texttt{Singularity}
file, a recipe akin to a \texttt{Dockerfile}, needs to be used to make
necessary adjustments. To date, no comparable image stack to the Rocker
Project's images exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which can
also build \texttt{Dockerfile}s and run Docker images. Proof of concepts
exists for using podman to build and run Rocker
containers\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}},
but the prevalence of Docker, especially in the broader user community
beyond experts or niche systems and the vast amount of blog posts and
courses for Docker currently cap specific development efforts for both
Singularity and podman in the R community. This might quickly change if
the usability and spread of Singularity or podman increase, or if
security features such as rootless/unprivileged containers, which both
these tools support out of the box, become more sought after.

@@

\begin{table}
\centering
\begin{tabular}{llllllll}
  \toprule
\rotatebox{-90}{Functionality} & \rotatebox{-90}{AzureContainers} & \rotatebox{-90}{babelwhale} & \rotatebox{-90}{dockermachine} & \rotatebox{-90}{dockyard} & \rotatebox{-90}{googleCloudRunner} & \rotatebox{-90}{harbor} & \rotatebox{-90}{stevedore }\\ 
  \midrule
Generate a Dockerfile &  &  &  & \checkmark &  &  &  \\ 
   \midrule
Build an image & \checkmark &  &  & \checkmark & \checkmark &  &  \\ 
   \midrule
Execute a container locally or remotely & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark & \checkmark \\ 
   \midrule
Deploy or manage instances in the cloud & \checkmark &  & \checkmark &  & \checkmark & \checkmark & \checkmark \\ 
   \midrule
Interact with an instance (e.g., file transfer) &  & \checkmark & \checkmark &  &  &  & \checkmark \\ 
   \midrule
Manage storage of images &  &  &  &  &  & \checkmark & \checkmark \\ 
   \midrule
Supports Docker and Singularity &  & \checkmark &  &  &  &  &  \\ 
   \midrule
Direct access to Docker API instead of using the CLI &  &  &  &  &  &  & \checkmark \\ 
   \midrule
Installing Docker software &  &  & \checkmark &  &  &  &  \\ 
   \bottomrule
\end{tabular}
\caption{R packages with Docker client functionality.} 
\label{tab:clients}
\end{table}

\textbf{\pkg{harbor}} (\url{https://github.com/wch/harbor}) is no longer
actively maintained, but it should be honourably mentioned as the first
R package for managing Docker images and containers. It uses the
\CRANpkg{sys} package \citep{cran_sys} to run system commands against
the Docker CLI, both locally and through an SSH connection, and it has
convenience functions, e.g., for listing and removing containers/images
and for accessing logs. The outputs of container executions are
converted to appropriate R types. The Docker CLI's basic functionality,
although it evolves quickly and with little concern for avoiding
breaking changes, has remained unchanged in core functions, meaning that
a core function such as
\code{harbor::docker\_run(image = "hello-world")} still works despite
its stopped development.

@@

\textbf{\pkg{dockermachine}}
(\url{https://github.com/cboettig/dockermachine}) is an R package to
provide a convenient interface to
\href{https://docs.docker.com/machine/overview/}{Docker~Machine} from R.
The CLI tool \texttt{docker-machine} allows users to create and manage a
virtual host on local computers, local data centres, or at cloud
providers. A local Docker installation can be configured to
transparently forward all commands issued on the local Docker CLI to a
selected (remote) virtual host. Docker~Machine was especially crucial
for local use in the early days of Docker, when no native support was
available for Mac or Windows computers, but it remains relevant for
provisioning on remote systems. The package has not received any updates
for two years, but it is functional with a current version of
\texttt{docker-machine} (\texttt{0.16.2}). It potentially lowers the
barriers for R users to run containers on various hosts if they perceive
that using the Docker~Machine CLI directly as a barrier and it enables
scripted workflows with remote processing.

@@

\textbf{Bioconductor} (\url{https://bioconductor.org/}) is an
open-source, open development project for the analysis and comprehension
of genomic data \citep{gentleman_bioconductor_2004}. As of October 30th
2019, the project consists of 1823 R software packages, as well as
packages containing annotation or experiment data. \emph{Bioconductor}
has a semi-annual release cycle, where each release is associated with a
particular version of R, and Docker images are provided for current and
past versions of \emph{Bioconductor} for convenience and
reproducibility. All images, which are described on the
\emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), are created with
\texttt{Dockerfile}s maintained on GitHub and distributed through
Docker~Hub\footnote{See \href{https://github.com/Bioconductor/bioconductor_docker}{https://github.com/Bioconductor/bioconductor\_docker} and \href{https://hub.docker.com/u/bioconductor}{https://hub.docker.com/u/bioconductor} respectively.}.
\emph{Bioconductor}'s ``base'' Docker images are built on top of the
\texttt{rocker/rstudio} image. \emph{Bioconductor} installs packages
based on the R version in combination with the Bioconductor version and,
therefore, uses Bioconductor version tagging \texttt{devel} and
\texttt{RELEASE\_X\_Y}, e.g., \texttt{RELEASE\_3\_10}. Past and current
combinations of R and \emph{Bioconductor} will therefore be accessible
via specific image tags.

The \emph{Bioconductor} \texttt{Dockerfile} selects the desired R
version from Rocker images, adds required system dependencies, and uses
the \CRANpkg{BiocManager} package for installing appropriate versions of
\emph{Bioconductor} packages \citep{cran_biocmanager}. A strength of
this approach is that the responsibility for complex software
configuration and customization is shifted from the user to the
experienced \emph{Bioconductor} core team. However, a recent audit of
the \emph{Bioconductor} image stack \texttt{Dockerfile} led to the
deprecation of several community-maintained images, because the numerous
specific images became too hard to understand, complex to maintain, and
cumbersome to customise. As part of the simplification, a recent
innovation is the \texttt{bioconductor\_docker:devel} image, which
emulates the \emph{Bioconductor} environment for nightly builds as
closely as possible. This image contains the environment variables and
the system dependencies needed to install and check almost all
\emph{Bioconductor} software packages (1813 out of 1823). It saves users
and package developers from creating this environment themselves.
Furthermore, the image is configured so that \code{.libPaths()} has
\samp{/usr/local/lib/R/host-site-library} as the first location. Users
mounting a location on the host file system to this location can
persistently manage installed packages across Docker containers or image
updates. Many R users pursue flexible workflows tailored to particular
analysis needs rather than standardized workflows. The new
\texttt{bioconductor\_docker} image is well suited for this preference,
while \texttt{bioconductor\_docker:devel} provides developers with a
test environment close to \emph{Bioconductor}'s build system.

\label{datascience} \textbf{Data science} is a widely discussed topic in
all academic disciplines \citep[e.g.,][]{donoho_50_2017}. These
discussions have shed light on the tools and craftspersonship behind the
analysis of data with computational methods. The practice of data
science often involves combining tools and software stacks and requires
a cross-cutting skillset. This complexity and an inherent concern for
openness and reproducibility in the data science community has led to
Docker being used widely. The remainder of this section presents example
Docker images and image stacks featuring R intended for data science.

@@

  the existing front-end configuration, but also to provide consistent

@@

  management\footnote{See \href{https://docs.gigantum.com/docs/environment-management}{https://docs.gigantum.com/docs/environment-management}}.

@@

\textbf{\pkg{containerit}}
(\url{https://github.com/o2r-project/containerit/}) attempts to take
this one step further and includes these tools to automatically create a
\texttt{Dockerfile} that can execute a given workflow
\citep{nust_containerit_2019}. \pkg{containerit} accepts an R object of
classes \code{"sessionInfo"} or \code{"session\_info"} as input and
provides helper functions to derive these from workflows, e.g., an R
script or R Markdown document, by analysing the session state at the end
of the workflow. It relies on the \pkg{sysreqs}
(\url{https://github.com/r-hub/sysreqs/}) package and it's mapping of
package system dependencies to platform-specific installation package
names\footnote{See \href{https://sysreqs.r-hub.io/}{https://sysreqs.r-hub.io/}.}.
\pkg{containerit} uses \pkg{stevedore} to streamline the user
interaction and improve the created \texttt{Dockerfile}s, e.g., by
running a container for the desired base image to extract the already
available R packages.

@@

\textbf{\CRANpkg{liftr}} \citep{liftr2019} aims to solve the problem of
persistent reproducible reporting in statistical computing based on the
R Markdown format \citep{xie2018}. The irreproducibility of authoring
environments can become an issue for collaborative documents and
large-scale platforms for processing documents. \pkg{liftr} makes the
dynamic R Markdown document the main and sole workflow control file and
the only file that needs to be shared between collaborators for
consistent environments, e.g.~demonstrated in the DockFlow project
(\url{https://dockflow.org}). It introduces new fields to the document
header, allowing users to manually declare the versioned dependencies
required for rendering the document. The package then generates a
\texttt{Dockerfile} from this metadata and provides a utility function
to render the document inside a Docker container, i.e.,
\code{render\_docker("foo.Rmd")}. An RStudio addin even allows
compilation of documents with the single push of a button.

System dependencies are the domain of Docker, but for a full description
of the computing environment, one must also manage the R version and the
R packages. R versions are available via the versioned Rocker image
stack \citep{RJ-2017-065}.
\href{https://github.com/ColinFay/ronline}{r-online} leverages these
images and provides an app for helping users to detect breaking changes
between different R versions and for historic exploration of R. With a
standalone NodeJS app or
\href{https://srv.colinfay.me/r-online}{r-online}, the user can compare
a piece of code run in two separate versions of R. Internally, r-online
opens one or two Docker instances with the given version of R based on
Rocker images, executes a given piece of code, and returns the result to
the user. Regarding R package management, this can be achieved with
MRAN, or with packages such as \CRANpkg{checkpoint}
\citep{cran_checkpoint} and \CRANpkg{renv} \citep{renv2019}, which can
naturally be applied within images and containers. For example,
\pkg{renv} helps users to manage the state of the R library in a
reproducible way, further providing isolation and portability. While
\pkg{renv} does not cover system dependencies, the \pkg{renv}-based
environment can be transferred into a container either by restoring the
environment based on the main configuration file \texttt{renv.lock} or
by storing the \pkg{renv}-cache on the host and not in the container
\citep{ushey_using_2019}. With both the system dependencies and R
packages consciously managed in a Docker image, users can start using
containers as the \emph{only} environment for their workflows, which
allows them to work independently of physical
computers\footnote{Allowing them to be digital "nomads", cf. J.~Bryan's \href{https://github.com/jennybc/docker-why}{https://github.com/jennybc/docker-why}.}
and to assert a specific degree of confidence in the stability of a
developed software
\citep[cf. \texttt{README.Rmd} in][]{marwick_research_2017}.

@@

lines of R

@@

run any container, the package \CRANpkg{googleCloudRunner}
\citep{cran:googleCloudRunner} demonstrates how R can be used as the
control language for one-time or batch processing jobs and scheduling of
